---
title: "PM 566 Midterm"
author: "Dana Gonzalez"
format: html
editor: visual
embed-resources: TRUE
theme: cosmo
---
# Read in CSV File

```{r}
data <- read.table("~/Downloads/AtlasPlusTableData.csv", header = TRUE, sep = ",")
```

# Introduction

Hepatitis C virus (HCV) infection is the most common bloodborne infection in the United States. In addition to this, national incidence rates for the disease have nearly doubled in the last decade. While the "baby-boomer" generation (those born between 1946 and 1964, and most likely to have received a blood transfusion before universal screening guidelines) has historically been identified as a key at-risk population, an increasing number of new cases have been reported in younger individuals.

In this assignment, I will visualize trends in HCV incidence rate for the age groups included in the dataset, and determine which group has had the steepest or most dramatic increase in new cases for the period of interest.

This dataset has 6 variables: "Indicator" ("Acute Viral Hepatitis C" for all observations), "Year", "Age.Group," Cases", "Rate.per.100.000" and "X" (denotes the type of rate, "Incidence Rate" across all observations).

# Methods

My chosen dataset comes from the CDC's National Center for HIV, Viral Hepatitis, STD, and Tuberculosis Prevention (NCHHSTP) site. This center's site provides access to AtlasPlus, an interactive tool that allows users, like myself, to download disease surveillance data from the last 20 years. Using this tool, I downloaded national age-specific incidence rates for Hepatitis C (HCV) from 2000 to 2022. 

In cleaning the data, I chose to first remove the "Indicator" and "X" columns, as they were consistent across all observations. I also made sure to check for missing values, although collecting data through AtlasPlus should have not included any of this type of data. Further, I renamed the remaining variables to be more concise and consistent. Finally, I gathered measures of central tendency and spread for my continous variables using the summary() function.

```{r}
# Remove "Indicator" and "X" Columns
library(dplyr)
data <- data %>%
  select(-Indicator, -X)

# Check for Missing Data
missing_data <- colSums(is.na(data))
missing_data

# Rename Variables
data <- data |>
  rename(
    year = Year,
    age_group = Age.Group,
    cases = Cases,
    incidence_rate = Rate.per.100.000)

# Variable Summaries
summary(data)
```

# Preliminary Data

### Load Libraries

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(RColorBrewer)
```


### Number of Observations per Year

```{r}
year_summary <- data |>
  group_by(year) |>
  summarise(observations = n())

colnames(year_summary) <- c("Year", "Number of Observations")
kable(year_summary,
      align = "cc") |>
  kable_styling(latex_options = c("hold_position"), full_width = TRUE) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "6cm")
```

### Average Yearly Incidence per 100,000 (All Age Groups)

```{r}
average_yearly_incidence <- data |>
  group_by(year) |>
  summarise(
    average_rate = mean(incidence_rate, na.rm = TRUE),
    sd_rate = sd(incidence_rate, na.rm = TRUE),
    .groups = "drop")

colnames(average_yearly_incidence) <- c("Year", "Average Incidence Rate", "Standard Deviation")
kable(average_yearly_incidence,
      align = "ccc") |>
  kable_styling(latex_options = c("hold_position"), full_width = TRUE) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "6cm")
```

### Average Incidence per 100,000 by Age Group (All Years)

```{r}
average_group_incidence <- data |>
  group_by(age_group) |>
  summarise(
    average_group = mean(incidence_rate, na.rm = TRUE),
    sd_group = sd(incidence_rate, na.rm = TRUE),
    .groups = "drop")

colnames(average_group_incidence) <- c("Age Group", "Average Incidence Rate", "Standard Deviation")
kable(average_group_incidence,
      align = "ccc") |>
  kable_styling(latex_options = c("hold_position"), full_width = TRUE) |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "6cm")
```

### Lineplot of Yearly Incidence Rates per 100,000 by Age Groups

```{r}
ggplot(data = data) +
  geom_line(mapping = aes(x = year, y = incidence_rate, color = age_group, group = age_group)) +
  labs(title = "Yearly Incidence Rates per 100,000 by Age Groups",
       x = "Year",
       y = "Incidence Rate per 100,000",
       color = "Age Group") +
  scale_color_brewer(palette = "RdPu") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Lineplot of Average Yearly Incidence per 100,000 (All Age Groups)

```{r}
ggplot(data = average_yearly_incidence) +
  geom_line(mapping = aes(x = Year, y = `Average Incidence Rate`), color = "hotpink") +
  labs(title = "Average Yearly Incidence per 100,000 (All Age Groups)",
       x = "Year",
       y = "Average Incidence per 100,000 (All Age Groups)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Lineplot of Average Incidence per 100,000 by Age Group (All Years)

```{r, fig.width = 10}
ggplot(average_group_incidence, aes(x = `Age Group`, y = `Average Incidence Rate`, group = 1)) +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  geom_point(size = 3, color = "hotpink", fill = "hotpink") +
  geom_text(aes(label = `Average Incidence Rate`), 
            vjust = 0.5,
            hjust = -0.2,
            size = 3) +
  labs(title = "Boxplot of Average Incidence per 100,000 by Age Group (All Years)",
       x = "Age Group", 
       y = "Average Incidence per 100,000 by Age Group (All Years)") +
  theme_minimal()
```

