---
title: PM 566 Lab 5
author: Dana Gonzalez
format: html
editor: visual
embed-resources: true
theme: cosmo
---

# Setup in R

```{r}
library(data.table)
library(dtplyr)
library (dplyr)
library(tidyverse)
```

```{r}
stations <- fread("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations[, USAF := as.integer(USAF)]

stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

stations <- stations[!is.na(USAF)]

stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```

```{r}
download.file(
  "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
  destfile = file.path("~", "Downloads", "met_all.gz"),
  method   = "libcurl",
  timeout  = 60
)

met <- data.table::fread(file.path("~", "Downloads", "met_all.gz"))
met <- as.data.frame(met)
```

```{r}
merge(
  x     = met,      
  y     = stations, 
  by.x  = "USAFID",
  by.y  = "USAF", 
  all.x = TRUE,      
  all.y = FALSE
  ) |> nrow()

stations <- stations[!duplicated(stations$USAF), ]

met <- merge(
  x     = met,      
  y     = stations, 
  by.x  = "USAFID",
  by.y  = "USAF", 
  all.x = TRUE,      
  all.y = FALSE
  )
head(met[, c('USAFID', 'WBAN.y', 'STATE.y')], n = 4)
```

## Question 1: Representative station for the US

What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the `quantile()` function. Do these three coincide?

```{r}
quantiles_temp <- quantile(met$temp, na.rm = TRUE)
quantiles_wind <- quantile(met$wind.sp, na.rm = TRUE)
quantiles_pressure <- quantile(met$atm.press, na.rm = TRUE)

quantiles_df <- data.frame(
  metric = rep(c("Temperature", "Wind Speed", "Pressure"), each = length(quantiles_temp)),
  quantile = rep(names(quantiles_temp), times = 3),
  value = c(quantiles_temp, quantiles_wind, quantiles_pressure))

quantiles_df
```

```{r}
median_temp <- met %>% filter(temp == 23.5)
median(median_temp$USAFID)

median_wind <- met %>% filter(wind.sp == 2.1)
median(median_wind$USAFID)

median_pressure <- met %>% filter(atm.press == 1014.1)
median(median_pressure$USAFID)
```

The median stations in the US for temperature, wind speed, and atmospheric pressure are as follows: 722171, 722860, and 723564, respectively.

## Question 2: Representative station per state

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}
state_median_temp <- met %>%
  group_by(STATE.x) %>%
  summarize(median_temp = median(temp, na.rm = TRUE), .groups = 'drop')

state_median_temp

state_median_temp_usafid <- met %>%
  inner_join(state_median_temp, by = "STATE.x") %>%
  filter(temp == median_temp) %>%
  group_by(STATE.x) %>%
  summarize(median_usafid = median(USAFID, na.rm = TRUE), .groups = 'drop')

state_median_temp_usafid
```

```{r}
state_median_wind <- met %>%
  group_by(STATE.x) %>%
  summarize(median_wind = median(wind.sp, na.rm = TRUE), .groups = 'drop')

state_median_wind

state_median_wind_usafid <- met %>%
  inner_join(state_median_wind, by = "STATE.x") %>%
  filter(wind.sp == median_wind) %>%
  group_by(STATE.x) %>%
  summarize(median_usafid = median(USAFID, na.rm = TRUE), .groups = 'drop')

state_median_wind_usafid
```


```{r}
state_median_pressure <- met %>%
  group_by(STATE.x) %>%
  summarize(median_pressure = median(atm.press, na.rm = TRUE), .groups = 'drop')

state_median_pressure

state_median_pressure_usafid <- met %>%
  inner_join(state_median_pressure, by = "STATE.x") %>%
  filter(atm.press == median_pressure) %>%
  group_by(STATE.x) %>%
  summarize(median_usafid = floor(median(USAFID, na.rm = TRUE)), .groups = 'drop')

state_median_pressure_usafid
```

## Question 3: In the middle?

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all \~100 points in the same figure, applying different colors for those identified in this question.

```{r}
stations <- (unique(met[,c("LAT.x","LON.x")]))
library(leaflet)
leaflet(data = met) |> 
  addProviderTiles('CartoDB.Positron') |> 
  addCircles(lat = ~met$LAT.x, lng = ~met$LON.x,
             opacity = 0.01, fillOpacity = 0.001, radius = 1, color = ~pal(met$USAFID))
```

Knit the doc and save it on GitHub.

## Question 4: Means of means

Using the `quantile()` function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable "average temperature level," which you'll need to create.

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria:

-   low: temp \< 20
-   Mid: temp \>= 20 and temp \< 25
-   High: temp \>= 25

```{r}

```

Once you are done with that, you can compute the following:

-   Number of entries (records),
-   Number of NA entries,
-   Number of stations,
-   Number of states included, and
-   Mean temperature, wind-speed, and atmospheric pressure.

All by the levels described before.

```{r}

```

Knit the document, commit your changes, and push them to GitHub. Once you're done, you can email me the link to your repository or, preferably, link to the [Lab 5 Issue](https://github.com/USCbiostats/PM566/issues/68) in your commit message, as follows:

``` bash
git commit -a -m "Finalizing lab 5 https://github.com/USCbiostats/PM566/issues/68"
```

This will let me know which version of your repository to look at for grading purposes.
